\textbf{Seasonal Dynamics of Epiphytic Microbial Communities on Marine Macrophyte Surfaces}

### 2021-02-23 submission/manuscript.tex

\newpage
\linenumbers
\sisetup{mode=text}
\setlength\parindent{24pt}

\hypertarget{summary}{%
\subsection{Summary}\label{summary}}

@@

Marine macrophytes (seagrasses and macroalgae) are important ecosystem
engineers forming close associations with microorganisms belonging to
all three domains of life (Egan \emph{et al.}, 2013; Tarquinio \emph{et
al.}, 2019). Microbes can live within macrophyte tissue as endophytes or
can form epiphytic communities on surfaces of leaves and thalli (Egan
\emph{et al.}, 2013; Hollants \emph{et al.}, 2013; Tarquinio \emph{et
al.}, 2019). Epiphytic and endophytic microbial communities exhibit a
close functional relationship with the macrophyte host. It was proposed
that this close relationship constitutes a holobiont, an integrated
community where the macrophyte organism and its symbiotic partners
support each other (Margulis, 1991; Egan \emph{et al.}, 2013; Tarquinio
\emph{et al.}, 2019).

Biofilms of microbial epiphytes can contain diverse taxonomic groups and
harbor cell abundances from 10\textsuperscript{2} to
10\textsuperscript{7} \si{\cells\per\cm\squared} (Armstrong \emph{et
al.}, 2000; Bengtsson \emph{et al.}, 2010; Burke and Thomas \emph{et
al.}, 2011). In such an environment a number of positive and negative
interactions between the macrophyte and the colonizing microorganisms
have been described (Egan \emph{et al.}, 2013; Hollants \emph{et al.},
2013; Tarquinio \emph{et al.}, 2019). Macrophytes can promote growth of
associated microbes by nutrient exudation, while in return
microorganisms may support macrophyte performance through improved
nutrient availability, phytohormone production and protection from toxic
compounds, oxidative stress, biofouling organisms and pathogens (Egan
\emph{et al.}, 2013; Hollants \emph{et al.}, 2013; Tarquinio \emph{et
al.}, 2019). Besides these positive interactions, macrophytes can
negatively impact the associated microbes such as pathogenic bacteria by
producing reactive oxygen species and secondary metabolites (Egan
\emph{et al.}, 2013; Hollants \emph{et al.}, 2013; Tarquinio \emph{et
al.}, 2019).

All these ecological roles are carried out by a taxonomically diverse
community of microorganisms. At higher taxonomic ranks a core set of
macrophyte epiphytic taxa was described consisting mainly of
\emph{Alphaproteobacteria}, \emph{Gammaproteobacteria},
\emph{Desulfobacterota}, \emph{Bacteroidota}, \emph{Cyanobacteria},
\emph{Actinobacteriota}, \emph{Firmicutes}, \emph{Planctomycetota},
\emph{Chloroflexi} and \emph{Verrucomicrobiota} (Crump and Koch, 2008;
Tujula \emph{et al.}, 2010; Lachnit \emph{et al.}, 2011). In contrast,
at lower taxonomic ranks host specific microbial communities were found
(Lachnit \emph{et al.}, 2011; Roth-Schulze \emph{et al.}, 2016).
Recently, it was shown that even different morphological niches within
the same alga had a higher influence on the composition of the bacterial
community than biogeography or environmental factors (Morrissey \emph{et
al.}, 2019). While the microbial community composition varies between
host species, metagenomic analyses revealed that the majority of the
microbial functions are conserved (Burke and Peter Steinberg \emph{et
al.}, 2011; Roth-Schulze \emph{et al.}, 2016). This discrepancy between
the microbial taxonomic and functional composition might be explained by
the lottery hypothesis. It postulates that an initial random
colonization step takes places from a set of functionally equivalent
taxonomic groups resulting in taxonomically different epiphytic
communities sharing a core set of functional genes (Burke and Peter
Steinberg \emph{et al.}, 2011; Roth-Schulze \emph{et al.}, 2016). In
addition, some of the variation in the reported data could be attributed
to different techniques used in these studies, such as different
protocols for epiphytic cell detachment and/or DNA isolation, as no
standard protocol has been yet established to study epiphytic
communities (Ugarelli \emph{et al.}, 2019; Korlević \emph{et al.},
submitted).

The majority of studies describing macrophyte epiphytic microbial
communities did not include possible seasonal changes (Crump and Koch,
2008; Lachnit \emph{et al.}, 2009; Burke and Thomas \emph{et al.}, 2011;
Roth-Schulze \emph{et al.}, 2016; Ugarelli \emph{et al.}, 2019). If
seasonal changes were taken into account, low temporal frequency,
applied methodologies and/or limited number of analysed host species did
not allow a high taxonomic resolution (Tujula \emph{et al.}, 2010;
Lachnit \emph{et al.}, 2011; Miranda \emph{et al.}, 2013; Michelou
\emph{et al.}, 2013; Mancuso \emph{et al.}, 2016). In the present study
we describe the seasonal dynamics of bacterial and archaeal communities
on the surfaces of the seagrass \emph{Cymodocea nodosa} and siphonous
macroalgae \emph{Caulerpa cylindracea} determined on a mostly monthly
scale. Bacterial and archaeal epiphytes were sampled in a meadow of
\emph{C. nodosa} invaded by the invasive \emph{C. cylindracea} and in a
locality of only \emph{C. cylindracea} located in the proximity of the
seagrass meadow. For comparison, the microbial community of the ambient
seawater was also characterized.

@@

Sequencing of the 16S rRNA V4 region yielded a total of 1.8 million
sequences after quality curation and exclusion of sequences without
known relatives (no relative sequences) and eukaryotic, chloroplast and
mitochondrial sequences (\autoref{supp-nseq_notus}). A total of 35
samples originating from epiphytic archaeal and bacterial communities
associated with surfaces of the seagrass \emph{C. nodosa} and the
macroalga \emph{C. cylindracea} were analysed. In addition, 18 samples
(one of the samples was sequenced twice) originating from the ambient
seawater were also processed for comparison. The number of reads per
sample ranged between 8,408 and 77,463 sequences
(\autoref{supp-nseq_notus}). Even when the highest sequencing effort was
applied the rarefaction curves did not level off as commonly observed in
high-throughput 16S rRNA amplicon sequencing approaches
(\autoref{supp-rarefaction}). Following quality curation and exclusion
of sequences as mentioned above reads were clustered into 28,750
different OTUs at a similarity level of 97 \si{\percent}. Read numbers
were normalized to the minimum number of sequences (8,408,
\autoref{supp-nseq_notus}) through rarefaction resulting in 17,079
different OTUs with 384 to 2,062 OTUs per sample
(\autoref{supp-calculators}). To determine seasonal changes in richness
and diversity the observed number of OTUs, Chao1, ACE, Exponential
Shannon and Inverse Simpson (Jost, 2006) were calculated after
normalization through rarefaction. Generally, richness estimators and
diversity indices showed similar trends. On average, higher values were
found for \emph{C. cylindracea} (mixed {[}Number of OTUs, 1,684.1 ±
133.8 OTUs{]} and monospecific {[}Number of OTUs, 1,731.9 ± 163.8
OTUs{]}) than for \emph{C. nodosa} (Number of OTUs, 1,058.2 ± 214.5
OTUs) and lowest values were obtained for the microbial community of the
ambient seawater (Number of OTUs, 523.9 ± 138.7 OTUs)
(\autoref{supp-calculators}). Seasonal changes did not reveal such large
dissimilarities. \emph{C. nodosa} communities showed a slow increase
towards the end of the study, while \emph{C. cylindracea} (mixed and
monospecific) communities were characterized by slightly higher values
in spring and summer than in autumn and winter
(\autoref{supp-calculators}).

To determine the proportion of shared archaeal and bacterial OTUs and
communities sampled in different environments the Jaccard's Similarity
Coefficient on presence-absence data and Bray-Curtis Similarity
Coefficient, respectively, were calculated. Coefficients were determined
after normalization through rarefaction and binning of samples from the
particular environment. The highest proportion of shared OTUs and
community was found between mixed and monospecific \emph{C. cylindracea}
(Jaccard, 0.35; Bray-Curtis, 0.78), while lower shared values were
calculated between ambient seawater and epiphytic communities
(\autoref{matrix}). Shared proportion between \emph{C. nodosa} and
\emph{C. cylindracea} were approximately in-between the values of mixed
and monospecific \emph{C. cylindracea}. To assess seasonal changes in
the proportion of shared OTUs and communities the Jaccard's and
Bray-Curtis Similarity Coefficients were calculated between consecutive
sampling points (\autoref{shared}). Both coefficients showed similar
trends. Temporal proportional changes were more pronounced for ambient
seawater than for \emph{C. nodosa} and especially \emph{C. cylindracea}
associated communities (\autoref{shared}). In addition, only 0.4 -- 1.0
\si{\percent} of OTUs from each surface associated community were
present at all seasons. These persistent OTUs constituted a high
proportion of total sequences (39.5 -- 52.3 \si{\percent}). To further
disentangle the environmental and seasonal community dissimilarity a
Principal Coordinates Analysis (PCoA) based on Bray-Curtis distances and
OTU abundances was applied. A clear separation between ambient seawater
and surface associated communities was found (\autoref{pcoa}). In
addition, a separation of epiphytic bacterial and archaeal communities
based on host species was detected. This separation was further
supported by ANOSIM (R = 0.96, \emph{p} \textless{} 0.001). Seasonal
changes of \emph{C. nodosa} associated communities indicated a
separation between spring, summer and autumn/winter samples (ANOSIM, R =
0.54, \emph{p} \textless{} 0.01). For \emph{C. cylindracea} associated
communities a separation between summer and autumn/winter/spring samples
was observed that was, however, not as strong as for \emph{C. nodosa}
associated communities (ANOSIM, R = 0.30, \emph{p} \textless{} 0.05)
(\autoref{pcoa}).

@@

Surfaces of marine macrophytes harbor biofilms consisting of diverse
microbial taxa (Egan \emph{et al.}, 2013; Tarquinio \emph{et al.},
2019). No standard protocol has been developed to study these
macophyte-associated microbes (Ugarelli \emph{et al.}, 2019). Different
procedures for removal of microbial cells from host surfaces are
described, such as host tissue shaking (Nõges \emph{et al.}, 2010),
scraping (Uku \emph{et al.}, 2007), swabbing (Mancuso \emph{et al.},
2016) and ultrasonication (Cai \emph{et al.}, 2014). All these methods
result in different removal efficiencies but none was enabling a
complete removal of attached microbial cells based on our experience. In
the present study, we applied a removal protocol (Korlević \emph{et
al.}, submitted) based on direct cellular lysis (Burke \emph{et al.},
2009). The application of a direct lysis procedure coupled with a high
sampling frequency and Illumina amplicon sequencing has enabled us to
described in detail the bacterial and archaeal communities associated
with the surfaces of two marine macrophytes, \emph{C. nodosa} and
\emph{C. cylindracea}.

In the present study, highest richness was observed for \emph{C.
cylindracea} (mixed and monospecific) followed by \emph{C. nodosa} and
lowest richness was found in ambient seawater microbial communities.
Higher richness of microbial communities associated with seagrasses than
in ambient seawater was described earlier and could be attributed to a
larger set of inhabitable microniches existing on macrophyte surfaces
than in the ambient seawater (Ugarelli \emph{et al.}, 2019). The highest
richness observed for \emph{C. cylindracea} might be partly due to its
contact with the sediment. The stolon of \emph{C. cylindracea} is
attached to the sediment surface with rhizoids and thus, the stolon and
rhizoids are in a direct contact with the sediment. In addition,
seasonal differences in richness observed for surface attached
communities indicated a slightly higher richness in spring and summer.
This pattern could be explained by a higher macrophyte growth in these
two seasons than in autumn and winter (M. Najdek, personal
communication; Zavodnik \emph{et al.}, 1998; Ruitton \emph{et al.},
2005). During their main growth season in spring and summer macrophytes
exhibit a more dynamic chemical interaction with the surface community
probably causing an increase in the number of inhabitable microniches
(Borges and Champenois, 2015; Rickert \emph{et al.}, 2016).

We observed a strong differentiation between the surface attached and
ambient seawater communities at the level of OTUs, in agreement with
most published studies (Burke and Thomas \emph{et al.}, 2011; Michelou
\emph{et al.}, 2013; Roth-Schulze \emph{et al.}, 2016; Mancuso \emph{et
al.}, 2016; Crump \emph{et al.}, 2018; Ugarelli \emph{et al.}, 2019).
This indicates that marine macrophytes are a selecting factor from the
pool of microbial taxa present in the ambient seawater, modifying the
microbial community once the macrophyte associated microbial biofilm
develops (Salaün \emph{et al.}, 2012; Michelou \emph{et al.}, 2013). In
contrast, Fahimipour \emph{et al.} (2017) report in a global study of
\emph{Zostera marina}, similarities between the microbial community
developed on leaves and in the ambient seawater. The discrepancy between
our data and the study of Fahimipour \emph{et al.} (2017) could be
explained by different seagrass species, methodological variations or
biogeographic trends as Fahimipour \emph{et al.} (2017) analysed samples
from different locations throughout the northern hemisphere. It is
possible that the microbial communities in ambient seawater and on
leaves from the same location are differing but are still more similar
to each other when compared to other sampling locations. Indeed, it was
found that prokaryotic communities vary substantially between different
sampling sites (Bengtsson \emph{et al.}, 2017). When the taxonomic
composition at high ranks was analysed no such strong differentiation
was noticed. Phyla and classes such as \emph{Actinobacteriota},
\emph{Bacteroidota}, \emph{Cyanobacteria}, \emph{Alphaproteobacteria},
\emph{Gammaproteobacteria} and \emph{Verrucomicrobiota} were found in
both ambient seawater as well as macrophyte associated, in agreement
with previous studies (Burke and Thomas \emph{et al.}, 2011; Egan
\emph{et al.}, 2013; Michelou \emph{et al.}, 2013). In contrast, when
low taxonomic ranks were analysed (i.e., family and genus) a strong
differentiation was observed (Figs. \ref{cyano} -- \ref{desulfo}). A
similar differentiation at lower microbial taxonomic ranks between
ambient seawater and macrophytes was described for other macrophyte
species as well (Egan \emph{et al.}, 2013; Michelou \emph{et al.}, 2013;
Ugarelli \emph{et al.}, 2019).

Beside differences between ambient seawater and surface associated
microbial communities, it is unclear whether the prokaryotic epiphytic
community is host-specific or whether there are generalist taxa
characteristic to all or many macrophytes (Egan \emph{et al.}, 2013).
Similar to previously described differences between microbial
communities in the ambient seawater and on macrophytes, at high
taxonomic ranks no major difference between the microbial communities
associated with different hosts was observed. The only high rank phylum
that was differing between \emph{C. nodosa} and \emph{C. cylindracea}
was \emph{Desulfobacterota}, with more abundant sequences in the
\emph{C. cylindracea} associated community. As already mentioned, the
rhizoids and part of the stolon are in contact with the sediment. Thus
\emph{Desulfobacterota} are probably a part of the epiphytic community
that was in contact with the sediment. Similar high rank taxa found in
this study were described to be specific for other species of
macrophytes (Burke and Thomas \emph{et al.}, 2011; Lachnit \emph{et
al.}, 2011; Mancuso \emph{et al.}, 2016; Bengtsson \emph{et al.}, 2017).
In contrast to high taxonomic ranks, a substantial differentiation
between host specific communities was found supporting the notion that
macrophyte associated microbial communities might be host-specific.
Host-specificity was also observed for other species of macroalgae and
seagrasses (Lachnit \emph{et al.}, 2011; Roth-Schulze \emph{et al.},
2016; Ugarelli \emph{et al.}, 2019; Morrissey \emph{et al.}, 2019).
Taken together, at high taxonomic ranks a core set of taxa could be
described that is characteristic for all or many macrophytes, while at
low taxonomic ranks a community specific to host species was identified
(Figs. \ref{pcoa} and \ref{community}) (Egan \emph{et al.}, 2013).

>>>>

\newpage
\linenumbers
\sisetup{mode=text}
\setlength\parindent{24pt}
\doublespacing

\hypertarget{abstract}{%
\subsection{Abstract}\label{abstract}}

@@

Marine macrophytes (seagrasses and macroalgae) are important ecosystem
engineers forming close associations with microorganisms belonging to
all three domains of life (Egan et al., 2013; Tarquinio et al., 2019).
Microbes can live within macrophyte tissue as endophytes or can form
epiphytic communities on surfaces of leaves and thalli (Egan et al.,
2013; Hollants et al., 2013; Tarquinio et al., 2019). Epiphytic and
endophytic microbial communities exhibit a close functional relationship
with the macrophyte host. It was proposed that this close relationship
constitutes a holobiont, an integrated community where the macrophyte
organism and its symbiotic partners support each other (Margulis, 1991;
Egan et al., 2013; Tarquinio et al., 2019).

Biofilms of microbial epiphytes can contain diverse taxonomic groups and
harbor cell abundances from 10\textsuperscript{2} to
10\textsuperscript{7} \si{\cells\per\cm\squared} (Armstrong et al.,
2000; Bengtsson et al., 2010; Burke et al., 2011b). In such an
environment a number of positive and negative interactions between the
macrophyte and the colonizing microorganisms have been described (Egan
et al., 2013; Hollants et al., 2013; Tarquinio et al., 2019).
Macrophytes can promote growth of associated microbes by nutrient
exudation, while in return microorganisms may support macrophyte
performance through improved nutrient availability, phytohormone
production and protection from toxic compounds, oxidative stress,
biofouling organisms and pathogens (Egan et al., 2013; Hollants et al.,
2013; Tarquinio et al., 2019). Besides these positive interactions,
macrophytes can negatively impact the associated microbes such as
pathogenic bacteria by producing reactive oxygen species and secondary
metabolites (Egan et al., 2013; Hollants et al., 2013; Tarquinio et al.,
2019).

All these ecological roles are carried out by a taxonomically diverse
community of microorganisms. At higher taxonomic ranks a core set of
macrophyte epiphytic taxa was described consisting mainly of
\emph{Alphaproteobacteria}, \emph{Gammaproteobacteria},
\emph{Desulfobacterota}, \emph{Bacteroidota}, \emph{Cyanobacteria},
\emph{Actinobacteriota}, \emph{Firmicutes}, \emph{Planctomycetota},
\emph{Chloroflexi} and \emph{Verrucomicrobiota} (Crump and Koch, 2008;
Tujula et al., 2010; Lachnit et al., 2011). In contrast, at lower
taxonomic ranks host specific microbial communities were found (Lachnit
et al., 2011; Roth-Schulze et al., 2016). Recently, it was shown that
even different morphological niches within the same alga had a higher
influence on the composition of the bacterial community than
biogeography or environmental factors (Morrissey et al., 2019). While
the microbial community composition varies between host species,
metagenomic analyses revealed that the majority of the microbial
functions are conserved (Burke et al., 2011a; Roth-Schulze et al.,
2016). This discrepancy between the microbial taxonomic and functional
composition might be explained by the lottery hypothesis. It postulates
that an initial random colonization step takes places from a set of
functionally equivalent taxonomic groups resulting in taxonomically
different epiphytic communities sharing a core set of functional genes
(Burke et al., 2011a; Roth-Schulze et al., 2016). In addition, some of
the variation in the reported data could be attributed to different
techniques used in these studies, such as different protocols for
epiphytic cell detachment and/or DNA isolation, as no standard protocol
has been yet established to study epiphytic communities (Ugarelli et
al., 2019; Korlević et al., submitted).

The majority of studies describing macrophyte epiphytic microbial
communities did not include possible seasonal changes (Crump and Koch,
2008; Lachnit et al., 2009; Burke et al., 2011b; Roth-Schulze et al.,
2016; Ugarelli et al., 2019). If seasonal changes were taken into
account, low temporal frequency, applied methodologies and/or limited
number of analysed host species did not allow a high taxonomic
resolution (Tujula et al., 2010; Lachnit et al., 2011; Michelou et al.,
2013; Miranda et al., 2013; Mancuso et al., 2016). In the present study
we describe the seasonal dynamics of bacterial and archaeal communities
on the surfaces of the seagrass \emph{Cymodocea nodosa} and siphonous
macroalgae \emph{Caulerpa cylindracea} determined on a mostly monthly
scale. Bacterial and archaeal epiphytes were sampled in a meadow of
\emph{C. nodosa} invaded by the invasive \emph{C. cylindracea} and in a
locality of only \emph{C. cylindracea} located in the proximity of the
seagrass meadow. For comparison, the microbial community of the ambient
seawater was also characterized.

\hypertarget{materials-and-methods}{%
\subsection{Materials and methods}\label{materials-and-methods}}

DNA from surfaces of \emph{C. nodosa} and \emph{C. cylindracea} was
isolated using a previously modified and adapted protocol that allows
for a selective epiphytic DNA isolation (Massana et al., 1997; Korlević
et al., submitted). Briefly, leaves and thalli are incubated in a lysis
buffer and treated with lysozyme and proteinase K. Following the
incubations, the mixture containing lysed epiphytic cells was separated
from the leaves and thalli and extracted using phenol-chloroform.
Finally, the extracted DNA was precipitated using isopropanol. DNA from
seawater picoplankton was extracted from 0.2 \si{\um} polycarbonate
filters according to Massana et al. (1997) with a slight modification.
Following the phenol-chloroform extraction, 1/10 of chilled 3
\si{\Molar} sodium acetate (pH 5.2) was added. DNA was precipitated by
adding 1 volume of chilled isopropanol, incubating the mixtures
overnight at \num{-20} \si{\degreeCelsius} and centrifuging at 20,000 ×
\emph{g} and 4 \si{\degreeCelsius} for 21 \si{\minute}. The pellet was
washed twice with 500 \si{\ul} of chilled 70 \si{\percent} ethanol and
centrifuged after each washing step at 20,000 × \emph{g} and 4
\si{\degreeCelsius} for 5 \si{\minute}. Dried pellets were re-suspended
in 50 -- 100 \si{\ul} of deionized water.

Illumina MiSeq sequencing of the V4 region of the 16S rRNA gene was
performed as described previously (Korlević et al., submitted). The V4
region of the 16S rRNA gene was amplified using a two-step PCR
procedure. In the first PCR, the 515F
(\(5'\)-GTGYCAGCMGCCGCGGTAA-\(3'\)) and 806R
(\(5'\)-GGACTACNVGGGTWTCTAAT-\(3'\)) primers from the Earth Microbiome
Project (\url{https://earthmicrobiome.org/protocols-and-standards/16s/})
were used (Caporaso et al., 2012; Apprill et al., 2015; Parada et al.,
2016). These primers contained on their \(5'\) end a tagged sequence.
Purified PCR products were sent for Illumina MiSeq sequencing at IMGM
Laboratories, Martinsried, Germany. Prior to sequencing at IMGM, the
second PCR amplification of the two-step PCR procedure was performed
using primers targeting the tagged region incorporated in the first PCR.
In addition, these primers contained adapter and sample-specific index
sequences. Beside samples, a positive and negative control for each
sequencing batch was sequenced. The negative control comprised PCR
reactions without DNA template, while for a positive control a mock
community composed of evenly mixed DNA material originating from 20
bacterial strains (ATCC MSA-1002, ATCC, USA) was used. Sequences
obtained in this study have been deposited in the European Nucleotide
Archive (ENA) at EMBL-EBI under accession number PRJEB37267.

Obtained sequences were analysed on the computer cluster Isabella
(University Computing Center, University of Zagreb) using mothur
(version 1.43.0) (Schloss et al., 2009) according to the MiSeq Standard
Operating Procedure (MiSeq SOP; \url{https://mothur.org/wiki/MiSeq_SOP})
(Kozich et al., 2013) and recommendations provided by the Riffomonas
project to enhance data reproducibility
(\url{http://www.riffomonas.org/}). For alignment and classification of
sequences the SILVA SSU Ref NR 99 database (release 138;
\url{http://www.arb-silva.de}) was used (Quast et al., 2013; Yilmaz et
al., 2014). Pipeline data processing and visualization was done using R
(version 3.6.0) (R Core Team, 2019) combined with packages vegan
(version 2.5-6) (Oksanen et al., 2019), tidyverse (version 1.3.0)
(Wickham et al., 2019) and multiple other packages ({\textbf{???}};
Neuwirth, 2014; Xie, 2014, 2015, 2019; Wilke, 2018; Xie et al., 2018;
Allaire et al., 2019; Zhu, 2019). The detailed analysis procedure
including the R Markdown file are available in the GitHub repository
(\url{https://github.com/MicrobesRovinj/Korlevic_EpiphyticDynamics_FrontMicrobiol_2021}).
Based on the ATCC MSA-1002 mock community included in the analysis an
average sequencing error rate of 0.01 \si{\percent} was determined,
which is in line with previously reported values for next-generation
sequencing data (Kozich et al., 2013; Schloss et al., 2016). In
addition, the negative controls processed together with the samples
yielded on average only 2 sequences after sequence quality curation.

@@

Sequencing of the 16S rRNA V4 region yielded a total of 1.8 million
sequences after quality curation and exclusion of sequences without
known relatives (no relative sequences) and eukaryotic, chloroplast and
mitochondrial sequences (\autoref{supp-nseq_notus}). A total of 35
samples originating from epiphytic archaeal and bacterial communities
associated with surfaces of the seagrass \emph{C. nodosa} and the
macroalga \emph{C. cylindracea} were analysed. In addition, 18 samples
(one of the samples was sequenced twice) originating from the ambient
seawater were also processed for comparison. The number of reads per
sample ranged between 8,408 and 77,463 sequences
(\autoref{supp-nseq_notus}). Even when the highest sequencing effort was
applied the rarefaction curves did not level off as commonly observed in
high-throughput 16S rRNA amplicon sequencing approaches
(\autoref{supp-rarefaction}). Following quality curation and exclusion
of sequences as mentioned above reads were clustered into 28,750
different OTUs at a similarity level of 97 \si{\percent}. Read numbers
were normalized to the minimum number of sequences (8,408,
\autoref{supp-nseq_notus}) through rarefaction resulting in 17,025
different OTUs with 347 to 1,977 OTUs per sample
(\autoref{supp-calculators}). To determine seasonal changes in richness
and diversity the observed number of OTUs, Chao1, ACE, Exponential
Shannon and Inverse Simpson (Jost, 2006) were calculated after
normalization through rarefaction. Generally, richness estimators and
diversity indices showed similar trends. On average, higher values were
found for \emph{C. cylindracea} (mixed {[}Number of OTUs, 1,683.5 ±
128.5 OTUs{]} and monospecific {[}Number of OTUs, 1,729.9 ± 145.8
OTUs{]}) than for \emph{C. nodosa} (Number of OTUs, 1,052.2 ± 211.4
OTUs) and lowest values were obtained for the microbial community of the
ambient seawater (Number of OTUs, 529.2 ± 146.8 OTUs)
(\autoref{supp-calculators}). Seasonal changes did not reveal such large
dissimilarities. \emph{C. nodosa} communities showed a slow increase
towards the end of the study, while \emph{C. cylindracea} (mixed and
monospecific) communities were characterized by slightly higher values
in spring and summer than in autumn and winter
(\autoref{supp-calculators}).

To determine the proportion of shared archaeal and bacterial OTUs and
communities sampled in different environments the Jaccard's Similarity
Coefficient on presence-absence data and Bray-Curtis Similarity
Coefficient, respectively, were calculated. Coefficients were determined
after normalization through rarefaction and binning of samples from the
particular environment. The highest proportion of shared OTUs and
community was found between mixed and monospecific \emph{C. cylindracea}
(Jaccard, 0.35; Bray-Curtis, 0.77), while lower shared values were
calculated between ambient seawater and epiphytic communities
(\autoref{matrix}). Shared proportion between \emph{C. nodosa} and
\emph{C. cylindracea} were approximately in-between the values of mixed
and monospecific \emph{C. cylindracea}. To assess seasonal changes in
the proportion of shared OTUs and communities the Jaccard's and
Bray-Curtis Similarity Coefficients were calculated between consecutive
sampling points (\autoref{shared}). Both coefficients showed similar
trends. Temporal proportional changes were more pronounced for ambient
seawater than for \emph{C. nodosa} and especially \emph{C. cylindracea}
associated communities (\autoref{shared}). In addition, only 0.4 -- 1.0
\si{\percent} of OTUs from each surface associated community were
present at all seasons. These persistent OTUs constituted a high
proportion of total sequences (40.2 -- 52.1 \si{\percent}). To further
disentangle the environmental and seasonal community dissimilarity a
Principal Coordinates Analysis (PCoA) based on Bray-Curtis distances and
OTU abundances was applied. A clear separation between ambient seawater
and surface associated communities was found (\autoref{pcoa}). In
addition, a separation of epiphytic bacterial and archaeal communities
based on host species was detected. This separation was further
supported by ANOSIM (R = 0.96, \emph{p} \textless{} 0.001). Seasonal
changes of \emph{C. nodosa} associated communities indicated a
separation between spring, summer and autumn/winter samples (ANOSIM, R =
0.54, \emph{p} \textless{} 0.01). For \emph{C. cylindracea} associated
communities a separation between summer and autumn/winter/spring samples
was observed that was, however, not as strong as for \emph{C. nodosa}
associated communities (ANOSIM, R = 0.30, \emph{p} \textless{} 0.05)
(\autoref{pcoa}).

@@

Surfaces of marine macrophytes harbor biofilms consisting of diverse
microbial taxa (Egan et al., 2013; Tarquinio et al., 2019). No standard
protocol has been developed to study these macophyte-associated microbes
(Ugarelli et al., 2019). Different procedures for removal of microbial
cells from host surfaces are described, such as host tissue shaking
(Nõges et al., 2010), scraping (Uku et al., 2007), swabbing (Mancuso et
al., 2016) and ultrasonication (Cai et al., 2014). All these methods
result in different removal efficiencies but none was enabling a
complete removal of attached microbial cells based on our experience. In
the present study, we applied a removal protocol (Korlević et al.,
submitted) based on direct cellular lysis (Burke et al., 2009). The
application of a direct lysis procedure coupled with a high sampling
frequency and Illumina amplicon sequencing has enabled us to described
in detail the bacterial and archaeal communities associated with the
surfaces of two marine macrophytes, \emph{C. nodosa} and \emph{C.
cylindracea}.

In the present study, highest richness was observed for \emph{C.
cylindracea} (mixed and monospecific) followed by \emph{C. nodosa} and
lowest richness was found in ambient seawater microbial communities.
Higher richness of microbial communities associated with seagrasses than
in ambient seawater was described earlier and could be attributed to a
larger set of inhabitable microniches existing on macrophyte surfaces
than in the ambient seawater (Ugarelli et al., 2019). The highest
richness observed for \emph{C. cylindracea} might be partly due to its
contact with the sediment. The stolon of \emph{C. cylindracea} is
attached to the sediment surface with rhizoids and thus, the stolon and
rhizoids are in a direct contact with the sediment. In addition,
seasonal differences in richness observed for surface attached
communities indicated a slightly higher richness in spring and summer.
This pattern could be explained by a higher macrophyte growth in these
two seasons than in autumn and winter (M. Najdek, personal
communication; Zavodnik et al., 1998; Ruitton et al., 2005). During
their main growth season in spring and summer macrophytes exhibit a more
dynamic chemical interaction with the surface community probably causing
an increase in the number of inhabitable microniches (Borges and
Champenois, 2015; Rickert et al., 2016).

We observed a strong differentiation between the surface attached and
ambient seawater communities at the level of OTUs, in agreement with
most published studies (Burke et al., 2011b; Michelou et al., 2013;
Mancuso et al., 2016; Roth-Schulze et al., 2016; Crump et al., 2018;
Ugarelli et al., 2019). This indicates that marine macrophytes are a
selecting factor from the pool of microbial taxa present in the ambient
seawater, modifying the microbial community once the macrophyte
associated microbial biofilm develops (Salaün et al., 2012; Michelou et
al., 2013). In contrast, Fahimipour et al. (2017) report in a global
study of \emph{Zostera marina}, similarities between the microbial
community developed on leaves and in the ambient seawater. The
discrepancy between our data and the study of Fahimipour et al. (2017)
could be explained by different seagrass species, methodological
variations or biogeographic trends as Fahimipour et al. (2017) analysed
samples from different locations throughout the northern hemisphere. It
is possible that the microbial communities in ambient seawater and on
leaves from the same location are differing but are still more similar
to each other when compared to other sampling locations. Indeed, it was
found that prokaryotic communities vary substantially between different
sampling sites (Bengtsson et al., 2017). When the taxonomic composition
at high ranks was analysed no such strong differentiation was noticed.
Phyla and classes such as \emph{Actinobacteriota}, \emph{Bacteroidota},
\emph{Cyanobacteria}, \emph{Alphaproteobacteria},
\emph{Gammaproteobacteria} and \emph{Verrucomicrobiota} were found in
both ambient seawater as well as macrophyte associated, in agreement
with previous studies (Burke et al., 2011b; Egan et al., 2013; Michelou
et al., 2013). In contrast, when low taxonomic ranks were analysed
(i.e., family and genus) a strong differentiation was observed (Figs.
\ref{cyano} -- \ref{desulfo}). A similar differentiation at lower
microbial taxonomic ranks between ambient seawater and macrophytes was
described for other macrophyte species as well (Egan et al., 2013;
Michelou et al., 2013; Ugarelli et al., 2019).

Beside differences between ambient seawater and surface associated
microbial communities, it is unclear whether the prokaryotic epiphytic
community is host-specific or whether there are generalist taxa
characteristic to all or many macrophytes (Egan et al., 2013). Similar
to previously described differences between microbial communities in the
ambient seawater and on macrophytes, at high taxonomic ranks no major
difference between the microbial communities associated with different
hosts was observed. The only high rank phylum that was differing between
\emph{C. nodosa} and \emph{C. cylindracea} was \emph{Desulfobacterota},
with more abundant sequences in the \emph{C. cylindracea} associated
community. As already mentioned, the rhizoids and part of the stolon are
in contact with the sediment. Thus \emph{Desulfobacterota} are probably
a part of the epiphytic community that was in contact with the sediment.
Similar high rank taxa found in this study were described to be specific
for other species of macrophytes (Burke et al., 2011b; Lachnit et al.,
2011; Mancuso et al., 2016; Bengtsson et al., 2017). In contrast to high
taxonomic ranks, a substantial differentiation between host specific
communities was found supporting the notion that macrophyte associated
microbial communities might be host-specific. Host-specificity was also
observed for other species of macroalgae and seagrasses (Lachnit et al.,
2011; Roth-Schulze et al., 2016; Morrissey et al., 2019; Ugarelli et
al., 2019). Taken together, at high taxonomic ranks a core set of taxa
could be described that is characteristic for all or many macrophytes,
while at low taxonomic ranks a community specific to host species was
identified (Figs. \ref{pcoa} and \ref{community}) (Egan et al., 2013).

Seasonal changes in richness in the epiphytic community were substantial
as indicated by the proportion of OTUs (\(\leq\) 1.0 \si{\percent})
present at every sampling date. These persistent OTUs, however, were
accounting for a high proportion of sequences (\(\geq\) 40.2
\si{\percent}) (\autoref{shared}). A very similar proportion of
persistent OTUs was reported in high-frequency sampling studies
describing seasonal changes in picoplankton (Gilbert et al., 2009,
2012). In comparison to the seawater community, a lower degree of
seasonal shifts was observed for the macrophyte surface associated
communities. It appears that microniches at the surfaces of macrophytes
are providing more stable conditions than the ambient seawater. At the
level of OTUs seasonal changes of \emph{C. nodosa} and \emph{C.
cylindracea} associated communities were identified that could be linked
to the growth cycle of the seagrass and macroalgae (M. Najdek, personal
communication). \emph{C. nodosa} was characterized by a spring community
during maximum seagrass proliferation, a summer community during the
highest standing stock of \emph{C.nodosa} and an autumn/winter community
during the decay of seagrass biomass. In contrast, \emph{C. cylindracea}
started to proliferate in late spring and was characterized only by a
summer community during high growth rates and by an autumn/winter/spring
community when the biomass was at the peak and decaying thereafter.
Similar seasonal changes in the epiphytic community were also described
for other macroalgae (Tujula et al., 2010; Lachnit et al., 2011). Higher
seasonal stability of \emph{C. cylindracea} surface communities than in
\emph{C. nodosa} was also observed in the higher proportion of shared
communities between two consecutive sampling dates in \emph{C.
cylindracea}.
